TADCompare: an R package for differential analysis and characterization of Topologically Associated Domains

### 2019-09-21 manuscript/manuscript.Rmd
Describe latest results sans p-values

We estimated coherence of 54 traits from five categories (Figure 3, Supplementary Table S3, Supplementary Figures S3-S5). Among anthropometric traits, "Weight" and "Body Mass Index" had the highest level of coherence. Other examples of traits with high coherence include blood pressure-related traits (cardiovascular) and glucose-related traits (metabolic). Notably, metabolic trait category had the highest overall coherence (mean coherence 0.91 and 1.05, respectively, Supplementary Table S1). Several notable examples of low coherence include "Obesity-related traits" and "Height" (anthropometric traits), and blood count-related traits (cardiovascular). Traits in the "Intelligence" category had low overall coherence (mean coherence = 0.76). These results highlight the relatively good knowledge of the genetics of metabolic traits which corresponds to highly coherent networks. Conversely, current knowledge about the genetics of intelligence-related traits appear insufficient as reflected in the low coherence of the corresponding networks.

@@

We further estimated coherence of 80 diseases from seven categories (Figure 4, Supplementary Figures S6-S8, Supplementary Table S3). Metabolic diseases had high overall coherence (average coherence 0.93), with "Obesity" and "Type 2 diabetes" being the most coherent (0.99 and 0.94, respectively). Similarly, the average coherence for neurologic diseases was 0.86, with "Alzheimer's disease" and "Parkinson's disease" being highly coherent (0.94 and 0.92, respectively). The high coherence of metabolic and neurologic disease networks highlight the relatively good knowledge of their genetics.

## Intermediate coherence of autoimmune, cancer, and eye diseases

"Rheumatoid arthritis", "Ankylosing spongylitis", "Psoriasis" were the most coherent in the "Autoimmune" category (0.94, 0.91, 0.89, respectively). On the other end of the coherence spectrum, "Multiple sclerosis" and "Allergic rinitis" were the least coherent (0.76 for both). The average coherence of autoimmune diseases was high, 0.85, with all but two autoimmune diseases having high coherence (> 0.80). Diseases in the "Cancer" category were similarly highly coherent (average coherence 0.84). "Non-small lung cancer", "Testicular germ cell tumor", "Epithelial ovarian cancer" were the most coherent (0.94, 0.91, 0.88, respectively) while "Basal cell carcinoma", "Breast cancer" "Hodgkin's lymphoma" were the least coherent (0.64, 0.78, 0.78, respectively). Overall, networks of all but "Basal cell carcinoma" cancers had high coherence (> 0.78). Diseases in the "Eye disease" category were more heterogeneous, with coherence ranging from 0.71 for "Myopia (pathological)" disease to 0.97 for "Refractive error". The average coherence of diseases in the "Eye disease" category was relatively high (0.86). These results suggest good overall knowledge of the genetics of autoimmune, cancer, and eye-related diseases and highlight limited knowledge for some diseases in these categories.

@@

Diseases in "Psychiatric" and "Cardiovascular" categories showed the lowest average level of coherence (0.82 and 0.83, respectively). The corresponding coherence estimates were highly heterogeneous. Among cardiovascular diseases, "Atrial fibrillation", "Coronary artery disease" showed the highest level of coherence (0.98 and 0.89, respectively), while "Sudden cardiac arrest" and "Hypertension" had the lowest coherence (0.70 and 0.77, respectively). Among psychiatric disorders, "Smoking behavior" and "Alcohol dependence" had high coherence (0.93 and 0.89, respectively). Notably, "Schizophrenia" and "Major depressive disorder" showed the lowest coherence not only in the "Psychiatric disease" category (0.67 and 0.75, respectively) but among all phenotypes. Other psychiatric disorders, such as "Bipolar disorder", "Autism", "ADHD", showed intermediate coherence estimates (0.87, 0.83, 0.78, respectively). In summary, these low coherence estimates highlight the need for further research on genetics of most cardiovascular and psychiatric diseases to better understand their molecular mechanisms, and suggests "Schizophrenia" and "Major depressive disorder" as the top candidates for further genetic studies.

@@

**Supplementary Table S1. Summary statistics of the analyzed phenotypes.** Minimum, mean, median, maximum of coherence estimates, the number of genes, SNPs, the total number of phenotype networks. Each worksheet corresponds to the PPI database used.

>>>>

We estimated coherence of 54 traits from five categories (Figure 3, Supplementary Table S3, Supplementary Figures S2-S3). Among anthropometric traits, "Weight" and "Body Mass Index" had the highest level of coherence. Other examples of traits with high coherence include blood pressure-related traits (cardiovascular) and glucose-related traits (metabolic). Notably, metabolic trait category had the highest overall coherence (mean coherence 0.91 and 1.05, respectively, Supplementary Table S1). Several notable examples of low coherence include "Obesity-related traits" and "Height" (anthropometric traits), and blood count-related traits (cardiovascular). Traits in the "Intelligence" category had low overall coherence (mean coherence = 0.76). These results highlight the relatively good knowledge of the genetics of metabolic traits which corresponds to highly coherent networks. Conversely, current knowledge about the genetics of intelligence-related traits appear insufficient as reflected in the low coherence of the corresponding networks.

@@

We further estimated coherence of 80 diseases from seven categories (Figure 4, Supplementary Figures S4-S5, Supplementary Table S3). Metabolic diseases had high overall coherence (average coherence 0.91), with "Obesity" and "Type 2 diabetes" being the most coherent (1.01 and 0.91, respectively). Similarly, the average coherence for cancer diseases was 0.72, with "Testicular germ cell tumor" and "Pancreatic cancer" being highly coherent (0.92 and 0.74, respectively). The high coherence of metabolic and cancer disease networks highlight the relatively good knowledge about their genetics.

## Intermediate coherence of autoimmune, neurologic, and eye diseases

"Rheumatoid arthritis", "Systemic Lupus Erythematosus", and "Systemic sclerosis" were the most coherent in the "Autoimmune" category (0.86, 0.80, 0.79, respectively). On the other end of the coherence spectrum, "Allergic rinitis" and "Atopic dermatitis" were the least coherent (0.51, 0.60, respectively). The average coherence of autoimmune diseases was high, 0.70 (Table 1), with all autoimmune diseases, except "Allergic rinitis", having coherence estimates > 0.60 (Supplementary Table S3). Diseases in the "Neurologic" category were similarly highly coherent (average coherence 0.72). However, the average coherence of neurologic diseases using STRING database was 0.49, suggesting that, with larger number of networks being estimated the overall coherence of neurologic diseases is low. "Alzheimer's disease" and "Parkinson's disease" were the most coherent (0.88 and 0.84, respectively), while "Migraine" was the least coherent (0.54). Diseases in the "Eye disease" category were more heterogeneous, with coherence ranging from 0.35 for "Corneal astigmatism" to 1.01 for "Refractive error". The average coherence of diseases in the "Eye disease" category was relatively high (0.66). These results suggest good overall knowledge of the genetics of autoimmune, neurologic, and eye-related diseases and highlight limited knowledge for some diseases in these categories.

@@

Diseases in "Psychiatric" and "Cardiovascular" categories showed the lowest average level of coherence (0.64 and 0.63, respectively). The corresponding coherence estimates were highly heterogeneous. Among cardiovascular diseases, "Coronary artery disease" showed the highest level of coherence (0.83), while "Hypertension" had the lowest coherence (0.53). Among psychiatric disorders, "Smoking behavior" and "Alcohol dependence" had high coherence (0.87 and 0.82, respectively). Notably, choerehce of "Schizophrenia" and "Major depressive disorder" networks (0.52 and 0.33, respectively) was at the lower end of coherence estimates among all phenotypes. Other psychiatric disorders, such as "Bipolar disorder", "ADHD", showed intermediate coherence estimates (0.76 and 0.55, respectively). In summary, these low coherence estimates highlight the need for further research on genetics of most cardiovascular and psychiatric diseases to better understand their molecular mechanisms. These results further suggest "Schizophrenia" and "Major depressive disorder" as the top candidates for further genetic studies in psychiatric genome-wide association studies.

@@

**Supplementary Table S1. Summary statistics of the analyzed phenotype categories.** Minimum, mean, median, maximum of coherence estimates, the number of genes, SNPs, the total number of phenotype networks. Each worksheet corresponds to the PPI database used.
